---
title: "Data access: Historical precipitation patterns and seasonal forecasts"
author: "Linda Petutschnig"
date: "25 5 2023"
output:
  html_document:
    theme: paper
    highlight: default
---

This script performs the following tasks:

## Historical precipitation data  
- Accesses the CHIRPS database and retrieves monthly precipitation data from 1990 to 2019.
- The data comes in tar.gz format and is unpacked before being cropped and masked to the area of interest.
- A raster brick is created from the 360 monthly layers and saved to hard drive.
- The tar.gz files and uncropped .tif files are deleted from hard drive. 

## Seasonal precipitation forecast  
- Accesses the ECMWF seasonal precipitation forecasts via the Copernicus Climate datastore.  
- A request is passed to the datastore and the result is returend and saved in grib format.  
- Writes the result to hard drive.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries

```{r load libraryies}

library(httr)
library(utils)
library(raster)
library(terra)
library(sf)
library(R.utils)
library(tidyr)
library(dplyr)
library(stringr)
library(here)

```

## Get historical data  

```{r get histrorical data}

start_date <- as.Date("1999-01-01")
end_date <- as.Date("2019-12-30")

# Define a list of years and months
date_range <- seq(start_date, end_date, by = "month")

urls <- lapply(date_range, function(date){
   sprintf("https://data.chc.ucsb.edu/products/CHIRPS-2.0/africa_monthly/tifs/chirps-v2.0.%s.%s.tif.gz", year = format(date, "%Y"), month = format(date, "%m"))
})

# Data download
download_data <- function(urls, dest_dir) {
  for (url in urls) {
    filename <- basename(url)
    dest_path <- file.path(dest_dir, filename)
    download.file(url, dest_path, quiet = FALSE)
  }
}

download_data(urls, here::here("data/rasters"))

```

## Unpack and stack data  

```{r process data}
## Process data
# set directory where tif.gz files are located
tif_dir <- here::here("data/rasters")
temp_dir <- tempdir()

aoi_shape <- st_read(here("data/geopackages","AOI_hex6.gpkg"))
# initialize list to store raster bricks

for (file in list.files(tif_dir, pattern = ".tif.gz", full.names = TRUE)) {
  # extract the file to a temporary directory
  gunzip(filename = file, temporary = TRUE, remove = TRUE)
    # get list of all files in the extracted directory
  extracted_files <- list.files(temp_dir, full.names = TRUE)
    # find the .tif file in the extracted directory
  tif_file <- extracted_files[grep(".tif", extracted_files)]
}

extent <- extent(aoi_shape)
# loop over each tar.gz file in the directory
prec_aoi_brick <-  terra::rast(tif_file) %>% 
  terra::crop(extent) %>% 
  terra::mask(aoi_shape)

# Executed
writeRaster(prec_aoi_brick, here::here("data/rasters","perc_aoi_brick_90_19.tif"), overwrite = TRUE)

# delete the extracted files
file.remove(extracted_files)
# delete the tar.gz files
file.remove(list.files(tif_dir, pattern = ".tar.gz", full.names = TRUE))

```
# Seasonal forecast data

```{r access ECMWF}
library(ecmwfr)

# To use this service, one must register as a user at the Copernicus climate Data Store and get an User ID and an API key
# To do so, follow instructions here (see section Use: Copernicus Climate Data Store (CDS)): https://github.com/bluegreen-labs/ecmwfr
# My personal credentials are stored in copernicus_credentials.R
source(here("copernicus_credentials.R"))

# Formulates a data request. The information can be copied from the Copernicus Climate Data Store.
# To do so, open a dataset to download and check all your required parameters. Then click on "Show API request".
request <- list(
 "dataset_short_name" = "seasonal-postprocessed-single-levels",
 "originating_centre" = "ecmwf",
 "system" = "5",
 "variable" = "total_precipitation_anomalous_rate_of_accumulation",
 "product_type" = "ensemble_mean",
 "year" = "2020",
 "month" = c("04","05"),
 "leadtime_month" = "1",
 "area" = "4.23/26.81/-5.02/35.0",
 "format" = "grib",
 "target" = "seasonal_prec_forecast_2020_0405_lead1.grib"
 )

# Actually downloads the data
file <- wf_request(
 user     = "36664",   # user ID (for authentification)
 request  = request,  # the request
 transfer = TRUE,     # download the file
 path     = "."       # store data in current working directory
 )

```