---
title: "Data analysis: Access to healthcare"
author: "Linda Petutschnig"
date: "13 6 2023"
output:
  html_document:
    theme: paper
    highlight: default
---

Two indicators are calculated that are both related to healthcare accessibility:  
  1: Average walking time to nearest healthcare facility  
  2: Number of people served by the same healthcare facility  
  
This script performs the following tasks:  

- Accesses the Malaria Atlas Project database and downloads the "Walking-only travel time to healthcare map without access to motorized transport" data.
- Plots the Accessibility raster to visualize the data  
 
- For indicator 1, the hexagonify function is used to calculate the average walking time in minutes per hexagon.
- For indicator 2, the population per service area is calculated and mapped. To this end, the "worldpop_2020_constrained_100m" data is loaded from hard drive which has been prepared in another script. The service areas were calculated using the SAGA GIS algorithm "Accumulate costs" with the "Walking-only travel time to healthcare map without access to motorized transport" data as input. The service areas are now overlaid with the population data. In a last step the population per service area is disaggregated into the hexagon layer.

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(ft.shadow = FALSE)

```

## Load libraries
```{r prepare, message = FALSE, warning = FALSE}

 # Loads necessary packages
library(here) # to access data via relative paths
library(malariaAtlas) # to access data provided by the Malaria Atlas Project
library(sf) # to manipultate vector features
library(raster) # to work with raster files

```

## Define extent and download data   
```{r accessibility get data, message = FALSE, warning = FALSE}

# Loads hexagon layer from database
aoi_shape = st_read(here("data/geopackages", "AOI_hex6.gpkg"))

# Converts AOI to SpatialPolygons object because this is required by the next function getRaster
aoi_spatial <- as(aoi_shape, "Spatial")

# Accesses the accessibility data via the malariaAtlas function "getRaster"
accessibility_raster <- getRaster(surface = "Walking-only travel time to healthcare map without access to motorized transport", shp = aoi_spatial)

# Plots the raw accessibility data 
plot(accessibility_raster)

# Saves raster layer to disk
writeRaster(accessibility_raster, here("data/rasters", "AOI_traveltime_foot_hf.tif"), overwrite = TRUE)

```

## Get population data  
```{r population get data, message = FALSE, warning = FALSE}

library(httr)
library(jsonlite)
library(raster)
library(sf)

# API path look up is in the following exemplified based on COD.
# Define the URL
url <- "https://hub.worldpop.org/rest/data/pop/wpicuadj1km?iso3=COD"

# Send the GET request
response <- GET(url)

# Parse the response data 
data <- content(response, "text")
parsed_data <- fromJSON(data) # Look into the parsed_data object for more information

#-----------------------------------
# Download country data

download_pop <- function(url, filename) {
  # Send GET request to the URL
  response <- GET(url)

  # Check if the request was successful (status code 200)
  if (status_code(response) == 200) {
    # File path where to save the downloaded .tif file
    file_path <- here("data/rasters", filename)

    # Save the content of the response to a file
    writeBin(content(response, "raw"), file_path)
    return(TRUE)
  } else {
    return(FALSE)
  }
}

cod_pop <- download_pop("https://data.worldpop.org/GIS/Population/Global_2000_2020_1km_UNadj/2020/COD/cod_ppp_2020_1km_Aggregated_UNadj.tif","COD_ppp_2020_unconst_1km_UNadj.tif")
bdi_pop <- download_pop("https://data.worldpop.org/GIS/Population/Global_2000_2020_1km_UNadj/2020/BDI/bdi_ppp_2020_1km_Aggregated_UNadj.tif","BDI_ppp_2020_unconst_1km_UNadj.tif")
rwa_pop <- download_pop("https://data.worldpop.org/GIS/Population/Global_2000_2020_1km_UNadj/2020/RWA/rwa_ppp_2020_1km_Aggregated_UNadj.tif","RWA_ppp_2020_unconst_1km_UNadj.tif")
uga_pop <- download_pop("https://data.worldpop.org/GIS/Population/Global_2000_2020_1km_UNadj/2020/UGA/uga_ppp_2020_1km_Aggregated_UNadj.tif","UGA_ppp_2020_unconst_1km_UNadj.tif")

#-----------------------------------
# Loads the just downloaded rasters.
pop_cod <- raster(here("data/rasters","COD_ppp_2020_unconst_1km_UNadj.tif"))
pop_bdi <- raster(here("data/rasters","BDI_ppp_2020_unconst_1km_UNadj.tif"))
pop_rwa <- raster(here("data/rasters","RWA_ppp_2020_unconst_1km_UNadj.tif"))
pop_uga <- raster(here("data/rasters","UGA_ppp_2020_unconst_1km_UNadj.tif"))

# To preserve the shape of the AOI, we first crop and then mask
AOI <- sf::read_sf(here("data/geopackages", "AOI.gpkg"))

crop_aoi <- function(country){
  pop_cod_aoi <- crop(country, extent(AOI)) %>% 
  mask(AOI)
}

# Crop and mask all population datasets to the boundary of the AOI
pop_cod_aoi <- crop_aoi(pop_cod)
pop_bdi_aoi <- crop_aoi(pop_bdi)
pop_rwa_aoi <- crop_aoi(pop_rwa)
pop_uga_aoi <- crop_aoi(pop_uga)

# Mosaic countries together and save
aoi_population <- mosaic(pop_cod_aoi, pop_bdi_aoi, pop_rwa_aoi, pop_uga_aoi, fun = "mean")
writeRaster(aoi_population, here("data/rasters","AOI_ppp_2020_unconst_UNadj.tif"))

## THINGS TRIED THAT DID NOT WORK----

# wopr-------
# Tried data access using wopr on 31.07.23. So far the catalog can only access Social Distancing and Building data for a few selected countries. Therefore, the package is not (yet)
# useful for accessing WorldPop data in general.
# library(wopr)
# catalogue <- getCatalogue()
# 

# wpgpDownloadR-------
# When using wpgpListCountryDatasets(), there is ppp_2020_UNadj_constrained in all four country lists. However, calling it receives an error, stating that the co-variate is not present in WP (?). Same goes for ppp_2020_UNadj.
# library(wpgpDownloadR)
# wpgpListCountries()
# wpgpListCountryDatasets(ISO3="COD")
# 
#  wpgpGetCountryDataset(ISO3 = "COD",
#                        covariate = "Unconstrained individual countries 2000-2020 UN adjusted ( 1km resolution )",
#                        destDir = here("data/rasters"))
# 

```

