---
title: "Armed conflicts locations"
author: "Linda Petutschnig"
date: "25 5 2023"
output:
  html_document:
    theme: paper
    highlight: default
---

This script performs the following tasks:

- Accesses the ACLED database and retrieves entries within a user-defined space and time frame.
- Loads a pre-prepared hexagonal DGGS grid.
- Applies a hotspot classification algorithm using the hexagonal grid as atomic units, categorizing the hexagons based on the number of ACLED events during a specified period.
- Converts the hotspot classes (no pattern, former hotspot, emerging hotspot, intermittent hotspot, persistent hotspot) into numeric values for better integration with other risk indicators.
- Writes the results to the hard drive and project database ("hotspots_3months").

By utilizing the ACLED database and the hexagonal DGGS grid, the script enables the identification of hotspots in armed conflict locations and population distribution. The generated hotspot classifications can be utilized for further analysis and risk assessment.
 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load, message = FALSE, warning = FALSE}

# Loads necessary packages
library(acled.api) # to access ACLED data
library(sf) # to manipulate vector features
library(dplyr) # for tidy data manipulation
library(here) # to access data via relative paths
library(sfhotspot) # to perform hotspot analysis based on vector features
library(tmap) # for map making

source(here("db-connection.R")) # to access data in the database
source(here("acled_credentials.R")) # contains the credentials necessary to access the ACLED data

# Add aceld.api
# https://www.rdocumentation.org/packages/acled.api/versions/1.1.5
# Citing ACLED: https://developer.acleddata.com/dashboard/terms-of-use/
 
```


```{r acled for aoi, message = FALSE, warning = FALSE}

# reads the hexagon shapes from the database
aoi_shape <- read_from_db("AOI_hex6")

# this function downloads ACLED data
get_acled <- function(country, from, to){
   v <- acled.api( # stores an ACLED sample in object v
     # EMAIL and ACCESS are sensitive information. 
     #Hence, they are stored in a separate file (acled_credentials.R), which is set as an additional source in the beginning of the script. 
      email.address = Sys.getenv("EMAIL_ADDRESS"),
      access.key = Sys.getenv("ACCESS_KEY"),
      country = country,
      start.date = from, 
      end.date = to,
      add.variables = c("longitude","latitude")
      )
 
      # convert into a spatial object (sf)
      v <- st_as_sf(v, coords=c("longitude","latitude"))
      
      # define the event dart´´tes as variables of type date
      v$event_date <- as.Date(v$event_date)
      
      # project
      st_crs(v) = 4326

   return(v)
}

# calls the above defined function and converts the results into an sf object
 acled_17_20_aoi <- get_acled(c("Democratic Republic of Congo","Rwanda","Uganda","Burundi"), "2017-01-01", "2020-12-31") %>% 
  st_as_sf() 

#  Writes the downloaded to hard drive
# st_write(acled_17_20_aoi, here("data","acled_17_20_aoi.shp"), append = TRUE)

```

```{r emerging hotspots, message = FALSE, warning = FALSE}

# calls the hotspot function on the just downloaded ACLED data
# the hexagon shapes are used as grid. For "period", 1,3,6 and 12 months were tested and visually compared. 
# 3 months were chosen for the analysis because it gave the most
# distinct pattern. Other periods may work as well. 
hotspots_3m <- hotspot_classify(acled_17_20_aoi, time = event_date, period = "3 months", grid = aoi_shape)

# plots the resulting hotspots
autoplot(hotspots_3m)

# writes a column with the hotspots classes into aoi_shape
aoi_shape$hotspot <- hotspots_3m$hotspot_category

# the hotspot categories are changed from categorical to nominal values to enable integration with the rest of the indicators. 
aoi_shape$hotspot_num[hotspots_3m$hotspot_category == 'no pattern'] <- "0"
aoi_shape$hotspot_num[hotspots_3m$hotspot_category == 'former hotspot'] <- "20"
aoi_shape$hotspot_num[hotspots_3m$hotspot_category == 'emerging hotspot'] <- "50"
aoi_shape$hotspot_num[hotspots_3m$hotspot_category == 'intermittent hotspot'] <- "80"
aoi_shape$hotspot_num[hotspots_3m$hotspot_category == 'persistent hotspot'] <- "100"

# converts the newly defined values to numeric format
aoi_shape$hotspot_num <- as.numeric(aoi_shape$hotspot_num)

#  writes the results to hard drive
# st_write(aoi_shape, here("data","acled_hotspot_3months.shp"), append = TRUE)

# writes results to the database
# write_to_db(aoi_shape, "hotspots_3months")

```


