---
title: "Long-term trends in malaria prevalence"
author: "Linda Petutschnig"
date: "28 04 2023"
output: 
  html_document:
    theme: paper
    highlight: default
---

This script performs the following tasks:

- Accesses the Malaria Atlas Project database and downloads the "Plasmodium falciparum incidence" data for all available years, cropping it to the Area of Interest (AOI).
- Manually adds additional years that were not available via direct download.
- Combines the individual years into a raster brick, which is then converted to points. Each point includes attribute columns with values for the respective years.
- Pre-processes and manipulates the data to prepare it for analysis.
- Generates a graph illustrating the development of Pf-incidence from 2000 to 2020 for all locations.
- Identifies a shift in the general trend around 2013 and conducts an in-depth analysis of the change between 2013 and 2020.
- Integrates the Pf-incidence values with a hexagonal grid prepared in a separate script.
- Uses the percentage change in Pf-incidence between 2013 and 2020 as a risk indicator and visualizes it on a map.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries

```{r malariaAtlas, message = FALSE, warning = FALSE}

library(malariaAtlas) # download malaria data
library(sf) # working with spatial vector data (simple features)
library(raster) # working with raster data
library(dplyr) # tidy data manipulation
library(here) # getting rid of relative paths
library(tidyr) # tidying data
library(lubridate) # to work with dates
library(stringr) # to manipulate sting variables
library(ggplot2) # plotting
library(tsibble) # time series analysis
library(fable) # time series analysis
library(MetBrewer) # color palettes for plots
library(tmap) # # for map making

source(here("db-connection.R")) # connect to database
source(here("hexagonify.R")) # a self-written function to summarize raster cells or points into hexagonal polygons

```

## Load and prepare data
```{r load PF incidence for aoi version 2020, error = FALSE, message = FALSE, warning = FALSE}

# reads the DGGS hexagons from database
aoi_shape <- read_from_db("AOI_hex6")

# getRatser needs a "SpatialPolygons" object as input, thus we convert the aoi here
aoi_spatial <- as(aoi_shape, "Spatial")

# this function downloads Plasmodium falciparum incidence from the MalariaAtlasProject site. We just need to specify the year.
pf <- function(year){
  getRaster( surface = "Plasmodium  falciparum  Incidence.", shp = aoi_spatial, year = year)
}

# download for 2000-2017
pf_00 <- pf(2000)
pf_01 <- pf(2001)
pf_02 <- pf(2002)
pf_03 <- pf(2003)
pf_04 <- pf(2004)
pf_05 <- pf(2005)
pf_06 <- pf(2006)
pf_07 <- pf(2007)
pf_08 <- pf(2008)
pf_09 <- pf(2009)
pf_10 <- pf(2010)
pf_11 <- pf(2011)
pf_12 <- pf(2012)
pf_13 <- pf(2013)
pf_14 <- pf(2014)
pf_15 <- pf(2015)
pf_16 <- pf(2016)
pf_17 <- pf(2017)

# for some reason, it is only possible to download the data between 2000 and 2017 via the getRaster function. 
# therefore, I downloaded 2018-2020 manually and it will now be read in, clipped and masked to our AOI and added to the malaria_cube.
pf_18 = raster(here("data","202206_Global_Pf_Incidence_Rate_2018.tif")) %>% 
  crop(aoi_shape) %>% 
  mask(pf_00)
pf_19 = raster(here("data","202206_Global_Pf_Incidence_Rate_2019.tif")) %>% 
  crop(aoi_shape) %>% 
  mask(pf_00)
pf_20 = raster(here("data","202206_Global_Pf_Incidence_Rate_2020.tif")) %>% 
  crop(aoi_shape) %>% 
  mask(pf_00)
  
# puts all rasters in one vector 
x <- c(pf_00, pf_01, pf_02, pf_03, pf_04, pf_05, pf_06, pf_07, pf_08, pf_09, pf_10, pf_11, pf_12, pf_13, pf_14, pf_15, pf_16, pf_17, pf_18, pf_19, pf_20)

# makes a 3-dimensional raster brick from all the malaria rasters from 2000 to 2020.
malaria_cube <- brick(x)

# convers the raster brick to points. Each point has the malaria values from 2000 to 2020 as variables. Then the points are transformed to sf objects for easier handling.
pf_00_20_points <- rasterToPoints(malaria_cube, spatial = TRUE) %>% 
  st_as_sf()

# Removes variables that we do not need anymore.
rm(aoi_spatial)

# writes the raster brick/cube to hard drive
# terra::writeRaster(malaria_cube,here("data/malaria_cube.tif"))

# writes the points dataset to hard drive
# sf::st_write(pf_00_20_points,here("data/pf_00_20_points.gpkg"))

# reads in the ready data 
# pf_00_20_points  <- st_read(here("data/pf_00_20_points.gpkg"))                        

```


## Prepare and inspect data

```{r data analysis with dplyr, message = FALSE, error = FALSE}

# adding an ID column to the points
pf_00_20_points <- pf_00_20_points %>% mutate(id = row_number())

# transforming the data into long format
# we want one Year column and a value column for our analysis.
# the gather function from the dplyr package is used to transform the data from wide to long format.
# (Note: consider using pivot_longer instead of gather, as it is the new function with the same functionality.)
pf_long <- pf_00_20_points %>% 
  gather(Year, MalariaIncidence, Plasmodium..falciparum..Incidence..2000:X202206_Global_Pf_Incidence_Rate_2020) 

# cleaning up the Year column
# the Year column currently has long strings like "Plasmodium..falciparum..Incidence..2000".
# we only need the last four characters, which represent the year. 
pf_long$Year <- str_sub(pf_long$Year,-4,-1)      

# converting the Year column into a proper date format
# we add Jan 1st to each row to create a proper date.
pf_long$Year <- lubridate::as_date(paste0(pf_long$Year, "-01-01"))

# converting the long format data table to a time series tibble by id
# every geographical point is represented in 21 rows between 2000 and 2020, but they share the same id.
pf_tsbl <- as_tsibble(pf_long, key = id, index = Year)

# calculating the average Malaria Incidence across all locations per year
avg <- pf_long %>%
  group_by(Year) %>% 
  summarise(avg_value = mean(MalariaIncidence, na.rm = TRUE))

# creating a time series plot per location as a line
ggplot() + 
  geom_line(data =  pf_long, aes(x = Year, y = MalariaIncidence, group =id), color = "darkseagreen", alpha = 0.1) +
  geom_line(data = avg, aes(x = Year, y = avg_value), color = "salmon", linewidth = 2) +
  labs(x = "Year", y = "Malaria Incidence", title = "Malaria Incidence Over Time") +
  theme_minimal(base_size = 14) 
  
```

## Filter relevant timespan
```{r since 2013, message = FALSE, error = FALSE, eval= FALSE}

# the overall malaria incidence has decreased drastically between 2000 and 2020. 
# however, the figure above reveals that there has been a shift in the trend: Since around 2013, numbers are rising again. 
# therefore, we remove everything before 2013 and just focus on the more recent years to calculate a trend.
pf_tsbl_recent <- pf_tsbl %>% 
  dplyr::filter(Year > "2012-01-01")

avg_recent <- avg %>% 
  dplyr::filter(Year > "2012-01-01") 

# this function calculates the percantage by which the malaria incidence value has changed from 2013 to 2020.
perc_avg_13 <- ((avg_recent$avg_value[8] - avg_recent$avg_value[1]) / avg_recent$avg_value[1]) * 100

# plot the data between 2013 and 2020
ggplot()+ 
  geom_line(data =  pf_tsbl_recent, aes(x = Year, y = MalariaIncidence, group =id), color = "darkseagreen", alpha = 0.1) +
  geom_line(data = avg_recent, aes(x= Year, y = avg_value), color = "salmon", linewidth = 2) +
  labs(x = "Year", y = "Malaria Incidence", title = "Malaria Incidence change since 2013") +
  theme_minimal(base_size = 14) 

# calculate the percentage of change in malaria incidence between 2013 and 2020 in every location.
# filter the values for 2013 and 2020
value_2000 <- pf_tsbl %>% 
  dplyr::filter(Year == "2000-01-01") %>% 
  dplyr::pull(MalariaIncidence)

value_2013 <- pf_tsbl %>% 
  dplyr::filter(Year == "2013-01-01") %>% 
  dplyr::pull(MalariaIncidence)

value_2020 <- pf_tsbl %>% 
  dplyr::filter(Year == "2020-01-01") %>% 
  dplyr::pull(MalariaIncidence)

# calculate the percentage.
perc_change_13 <- ((value_2020 - value_2013) / value_2013) * 100
perc_change_00 <- ((value_2020 - value_2000) / value_2000) * 100

# write the percentage change in malaria incidence into a new column that we add to the spatial dataset pf_00_20_points
pf_00_20_points$perc_change_13 <- perc_change_13
pf_00_20_points$perc_change_00 <- perc_change_00

```

## Hexagonify
```{r hexa, message = FALSE, error = FALSE, eval= FALSE}

# the data that goes into the hexagonify_points function is only allowed to have one column with values (in addition to the geometry column).
# thus, we filter the column with the malaria incidence values first.
pf_recent_points <- pf_00_20_points[24]


hexa_points <- function(points, polygons, fun){
  
  # Checks which points intersect with which hexagon (polygons)
  poly_point_intersect <- st_intersects(polygons, points)
  
  # A function that is applied to every point per polygon, it returns the mean of all point values in the polygons 
  hexagon_value <- sapply(poly_point_intersect, function(ids) {
    pts <- points [ids, ]
    return(fun(pts[[1]])) # Accesses the value in the first column
  })
  
  # Fills in the average walking time per hexagon cell 
  return(hexagon_value)
}

aoi_shape$pf_hexa = hexa_points(pf_recent_points, aoi_shape, mean)

aoi_shape$pf_hexa[aoi_shape$pf_hexa == 0] <- NA

# writes the results to the database
# write_to_db(aoi_shape, "Indicators")

```

## Visualize the results in a map
```{r where, message = FALSE, error = FALSE, eval= FALSE}

# sets the tmap mode to an interactive map
tmap_mode("view")

# defines the color palette
pal = met.brewer("Gauguin", type = "continuous", direction = -1)

# creates the map 
# map suddenly given an error - why?
tm_shape(aoi_shape) +
  tm_polygons(col = "pf_hexa", palette = pal, popup.vars = c("Change 13-20: "="pf_hexa", id = "h3_index")) +
  tm_layout(title = "Change in malaria incidence between 2013 and 2020") + 
  tm_fill(col = '#BFE1F4') + 
  tm_borders(lwd = 1, col = "black")  

```


